---
title: "R Notebook - Bigger Bad Bugs in the Big City: Effects of Urbanization on Invasive Spotted Lanternfly (Lycorma delicatula) Body Size - Levine et al."

output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

First, load all necessary packages.
```{r}
#load packages
library(tidyverse)
library(hexbin)
library(patchwork)
library(RSQLite)
library(lubridate)
library(lme4)
library(DHARMa)
library(car)
library(HLMdiag)
library(MuMIn)
library(lmerTest)

```

Then, import and format the data. 
```{r}
#read data from csv
sizes <- read_csv("/Users/brennalevine/Desktop/SLF_body_size_ICB_ms/Levine_et_al_body_size.csv")

#declare factors
sizes$Sex <- factor(sizes$Sex)
sizes$Bag_Code <- factor(sizes$Bag_Code)
sizes$Measurer_Initials <- factor(sizes$Measurer_Initials)
sizes$Site_Code <- factor(sizes$Site_Code)

#omit NAs from data set
sizes <- na.omit(sizes)

#add new column calculating ratio of nose-abd to nose-wing
sizes <- sizes %>%
  mutate(length_ratio = `Nose-to-Abdomen-Tip-(mm)`/`Nose-To-Wing-Tip-(mm)`)

#make new values numeric
sizes$length_ratio <- as.numeric(sizes$length_ratio)

#make sure nose-to-wingtip and %IP are numeric
sizes$`Nose-To-Wing-Tip-(mm)` <- as.numeric(sizes$`Nose-To-Wing-Tip-(mm)`)
sizes$`%IP-5KM` <- as.numeric(sizes$`%IP-5KM`)

#use lubridate to format dates
sizes$coll_date <- as.numeric(ymd(paste("2021", sizes$Month, sizes$Day, sep = "-")))

#subset to male data set
males <- sizes %>%
  filter(Sex == 'M') %>%
  select(Measurer_Initials, Site_Code, `Nose-To-Wing-Tip-(mm)`, `Nose-to-Abdomen-Tip-(mm)`, Latitude, length_ratio, coll_date, `%IP-5KM`, Bag_Code, Num_TOH, Altitude)

#add columns for transformed outcome variables
males$`Nose-To-Wing-Tip-(mm)-LOG` <- log10(males$`Nose-To-Wing-Tip-(mm)`)
males$`Nose-to-Abdomen-Tip-(mm)-LOG` <- log10(males$`Nose-to-Abdomen-Tip-(mm)`)
males$`Nose-To-Wing-Tip-(mm)-SQRT` <- sqrt(males$`Nose-To-Wing-Tip-(mm)`)
males$`Nose-to-Abdomen-Tip-(mm)-SQRT` <- sqrt(males$`Nose-to-Abdomen-Tip-(mm)`)

#subset to a female data set
females <- sizes %>%
  filter(Sex == 'F') %>%
  select(Measurer_Initials, Site_Code, `Nose-To-Wing-Tip-(mm)`, `Nose-to-Abdomen-Tip-(mm)`, Latitude, length_ratio, coll_date, `%IP-5KM`, Bag_Code, Num_TOH, Altitude)

#add columns for transformed outcome variables
females$`Nose-To-Wing-Tip-(mm)-LOG` <- log10(females$`Nose-To-Wing-Tip-(mm)`)
females$`Nose-to-Abdomen-Tip-(mm)-LOG` <- log10(females$`Nose-to-Abdomen-Tip-(mm)`)
females$`Nose-To-Wing-Tip-(mm)-SQRT` <- sqrt(females$`Nose-To-Wing-Tip-(mm)`)
females$`Nose-to-Abdomen-Tip-(mm)-SQRT` <- sqrt(females$`Nose-to-Abdomen-Tip-(mm)`)

```
Scale variables for males and females
```{r}
#add columns for scaled variables - MALES
males$scaled_5KM <- as.numeric(scale(males$`%IP-5KM`, center = TRUE, scale = TRUE))
males$scaled_NW <- as.numeric(scale(males$`Nose-To-Wing-Tip-(mm)`, center = TRUE, scale = TRUE))
males$scaled_NA <- as.numeric(scale(males$`Nose-to-Abdomen-Tip-(mm)`, center = TRUE, scale = TRUE))
males$scaled_LR <- as.numeric(scale(males$length_ratio, center = TRUE, scale = TRUE))
males$scaledLAT <- as.numeric(scale(males$Latitude, center = TRUE, scale = TRUE))
males$scaledLR <- as.numeric(scale(males$length_ratio, center = TRUE, scale = TRUE))
males$coll_date_scaled <- as.numeric(scale(males$coll_date, center = TRUE, scale = TRUE))
males$Num_TOH_scaled <- as.numeric(scale(males$Num_TOH, center = TRUE, scale = TRUE))
males$Altitude_scaled <- as.numeric(scale(males$Altitude, center = TRUE, scale = TRUE))


#add columns for scaled variables - FEMALES
females$scaled_5KM <- as.numeric(scale(females$`%IP-5KM`, center = TRUE, scale = TRUE))
females$scaled_NW <- as.numeric(scale(females$`Nose-To-Wing-Tip-(mm)`, center = TRUE, scale = TRUE))
females$scaled_NA <- as.numeric(scale(females$`Nose-to-Abdomen-Tip-(mm)`, center = TRUE, scale = TRUE))
females$scaled_LR <- as.numeric(scale(females$length_ratio, center = TRUE, scale = TRUE))
females$scaledLAT <- as.numeric(scale(females$Latitude, center = TRUE, scale = TRUE))
females$scaledLR <- as.numeric(scale(females$length_ratio, center = TRUE, scale = TRUE))
females$coll_date_scaled <- as.numeric(scale(females$coll_date, center = TRUE, scale = TRUE))
females$Num_TOH_scaled <- as.numeric(scale(females$Num_TOH, center = TRUE, scale= TRUE))
females$Altitude_scaled <- as.numeric(scale(females$Altitude, center = TRUE, scale = TRUE))

```

Perform data visualization.
```{r}
plot_sex_sizes <- ggplot(data = sizes, mapping = aes(x = Sex, y = `Nose-To-Wing-Tip-(mm)`, color = Sex)) +
  geom_boxplot(color = "black") +
  geom_jitter(alpha = 0.3) +
  scale_color_manual(values = c("F" = "red", "M" = "blue")) +
  guides(color = "none") +
  labs(x = "Sex", y = "Body Length (mm)") +
  theme_classic() +
  theme(panel.grid = element_blank())

# Save the plot as a 600 DPI PNG file
ggsave("plot_sex_sizes_600dpi.png", plot = plot_sex_sizes, dpi = 600, width = 8, height = 6, units = "in")

# Calculate mean and standard deviation for each sex
summary_stats <- sizes %>%
  group_by(Sex) %>%
  summarise(
    mean_length = mean(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE),
    sd_length = sd(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE)
  )

print(summary_stats)



```
#Evaluate correlation between altitude and urbanization.
```{r}
# Correlation between altitude and urbanization - overall
corr_altitude_urbanization <- cor.test(sizes$Altitude, sizes$`%IP-5KM`)
print(corr_altitude_urbanization)

# Correlation between altitude and urbanization - MALES AND FEMALES
male_corr_altitude_urbanization <- cor.test(males$Altitude, males$`%IP-5KM`)
female_corr_altitude_urbanization <- cor.test(females$Altitude, females$`%IP-5KM`)

print(male_corr_altitude_urbanization)
print(female_corr_altitude_urbanization)

#evaluate correlation between altitude/urbanization and nose-to-wingtip length - MALES 
male_corr_altitude_bodylength <- cor.test(males$Altitude, males$`Nose-To-Wing-Tip-(mm)`)
male_corr_urbanization_bodylength <- cor.test(males$`%IP-5KM`, males$`Nose-To-Wing-Tip-(mm)`)
                                              
print(male_corr_altitude_bodylength)
print(male_corr_urbanization_bodylength)

#evaluate correlation between altitude/urbanization and nose-to-wingtip length for FEMALES
female_corr_altitude_bodylength <- cor.test(females$Altitude, females$`Nose-To-Wing-Tip-(mm)`)
female_corr_urbanization_bodylength <- cor.test(females$`%IP-5KM`, females$`Nose-To-Wing-Tip-(mm)`)
                                              
print(female_corr_altitude_bodylength)
print(female_corr_urbanization_bodylength)

```

#Compare male and female body length
```{r}
# T-test for body length
length_ttest <- t.test(females$`Nose-To-Wing-Tip-(mm)`, males$`Nose-To-Wing-Tip-(mm)`, var.equal = FALSE)
print(length_ttest)

# Calculate mean and standard deviation for body length for males
male_summary <- males %>%
  summarise(
    Mean_Length_M = mean(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE),
    SD_Length_M = sd(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE),
  )

# Calculate mean and standard deviation for weight and Nose-to-Wingtip length for females
female_summary <- females %>%
  summarise(
    Mean_Length_F = mean(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE),
    SD_Length_F = sd(`Nose-To-Wing-Tip-(mm)`, na.rm = TRUE),
  )

print(male_summary)
print(female_summary)

```

Data Visualization for body length.
```{r}
###### males
# Plot the relationship with a scatter plot
ggplot(males, aes(x = `%IP-5KM`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(x = "% Impervious Surface", y = "Body Length (mm)",
       title = "Males: % Impervious Surface vs. Body Length") +
  theme_minimal()

# Plot the relationship with a scatter plot
ggplot(males, aes(x = `Latitude`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(x = "Latitude", y = "Body Length (mm)",
       title = "Males: Latitude vs. Body Length") +
  theme_minimal()

#males
# Plot the relationship with a scatter plot
ggplot(males, aes(x = `Altitude`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(x = "Altitude", y = "Body Length (mm)",
       title = "Males: Altitude vs. Body Length") +
  theme_minimal()


#females
# Plot the relationship with a scatter plot
ggplot(females, aes(x = `%IP-5KM`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(x = "% Impervious Surface", y = "Body Length (mm)",
       title = "Females: % Impervious Surface vs. Body Length") +
  theme_minimal()

#females
# Plot the relationship with a scatter plot
ggplot(females, aes(x = `Latitude`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(x = "Latitude", y = "Body Length (mm)",
       title = "Females: Latitude vs. Body Length") +
  theme_minimal()

#females
# Plot the relationship with a scatter plot
ggplot(females, aes(x = `Altitude`, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(x = "Altitude", y = "Body Length (mm)",
       title = "Females: Altitude vs. Body Length") +
  theme_minimal()

```


Run linear models evaluating the effects of percent impervious surface and latitude on male body length. Use site code as a random effect. Include quadratic effects for %IP and Latitude as relationship between these variables and the dependent variable looks a little humped and regression diagnostics confirm. Latitude and percent impervious surface are scaled. 
```{r}

# Ensure the na.action is set to na.fail
options(na.action = na.fail)

#global model for 5 KM buffer zone
M_nose_wing_5KM <- lmer(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2) + (1|Site_Code), data=males, REML = FALSE)

#view models
summary(M_nose_wing_5KM)

```

Model selection and averaging using MuMIn.
```{r}
# Perform model selection using the dredge function
dredge_results_M_IP <- dredge(M_nose_wing_5KM)

# View the model selection table
dredge_results_M_IP

# Select the top models (e.g., with delta AICc < 2)
top_models_M_IP <- get.models(dredge_results_M_IP, subset = delta < 2)

# Perform model averaging
model_average_M_IP <- model.avg(top_models_M_IP)

# Summary of the model averaging results
summary(model_average_M_IP)
```

Check for collinearity of predictors using car.
```{r}
#extract fixed effects from model
vif_results_M_IP <- vif(lm(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2), data = males))
print(vif_results_M_IP)
```

Test model assumptions for best models. The global model was the best fit after averaging the models.
```{r}
#make DHarma qqplot
M_best_model_resid_5KM <- simulateResiduals(M_nose_wing_5KM, plot = T, n = 10000)

#view DHarma qqplot nplot 
M_best_model_resid_5KM

#check residuals against predictors
plotResiduals(M_best_model_resid_5KM)
plotResiduals(M_best_model_resid_5KM, form = males$scaled_5KM)
plotResiduals(M_best_model_resid_5KM, form = males$scaledLAT)

#test dispersion
testDispersion(M_best_model_resid_5KM)
testOutliers(M_best_model_resid_5KM, alternative = 'two.sided')
outliers(M_best_model_resid_5KM)

```

Re-analyze without outliers to see how it affects results.

First, remove the rows that contain the outliers.
```{r}
# Define the rows to remove
rows_to_remove <- c(78, 227, 317, 444, 554)
# Create the new data frame without the specified rows
males_no_outliers <- males[-rows_to_remove, ]
```

Re-run LMMs.
```{r}
# Ensure the na.action is set to na.fail
options(na.action = na.fail)

#global model for 5 KM buffer zone
M_nose_wing_5KM_no_outliers <- lmer(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2) + (1|Site_Code), data=males_no_outliers, REML = FALSE)

#view model
summary(M_nose_wing_5KM_no_outliers)
```


# ----------------------------------------------------------------------
# ---------------------------------------------------------------------

#Analysis of Female SLF

Run linear models evaluating the effects of percent impervious surface and latitude on female body length. Use site code as a random effect. Include quadratic effects for %IP and Latitude as relationship between these variables and the dependent variable looks a little humped and the original DHarma runs indicated missing quadratic terms. Latitude and percent impervious surface are scaled. 
```{r}

# Ensure the na.action is set to na.fail
options(na.action = na.fail)

#global model for 5 KM buffer zone
F_nose_wing_5KM <- lmer(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2) + (1 | Site_Code), data=females, REML = FALSE)

#view models
summary(F_nose_wing_5KM)

```

Model selection and averaging using MuMIn.
```{r}
# Perform model selection using the dredge function
dredge_results_F_IP <- dredge(F_nose_wing_5KM)

# View the model selection table
dredge_results_F_IP

# Select the top models (e.g., with delta AICc < 2)
top_models_F_IP <- get.models(dredge_results_F_IP, subset = delta < 2)

# Perform model averaging
model_average_F_IP <- model.avg(top_models_F_IP)

# Summary of the model averaging results
summary(model_average_F_IP)
```


```{r}

#make DHarma qqplot
F_best_model_resid_5KM <- simulateResiduals(F_nose_wing_5KM, plot = T, n = 20000)
#view DHarma qqplot nplot 

#check residuals against predictors
plotResiduals(F_best_model_resid_5KM)
plotResiduals(F_best_model_resid_5KM, form = females$scaled_5KM)
plotResiduals(F_best_model_resid_5KM, form = females$scaledLAT)
#test dispersion
testDispersion(F_best_model_resid_5KM)
testOutliers(F_best_model_resid_5KM, alternative = 'two.sided')
outliers(F_best_model_resid_5KM)


```


Check for collinearity of predictors using car.
```{r}
#extract fixed effects from model
vif_results_F_IP <- vif(lm(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2), data = females))
print(vif_results_F_IP)
```

Re-analyze without outliers to see how it affects results.

First, remove the rows that contain the outliers.
```{r}
# Define the rows to remove
rows_to_remove_F <- c(294, 1059, 1069)
# Create the new data frame without the specified rows
females_no_outliers <- females[-rows_to_remove_F, ]

```

```{r}

# Ensure the na.action is set to na.fail
options(na.action = na.fail)

#global model for 5 KM buffer zone
F_nose_wing_5KM_no_outliers <- lmer(`Nose-To-Wing-Tip-(mm)` ~ scaledLAT + I(scaledLAT^2) + scaled_5KM + I(scaled_5KM^2) + (1 | Site_Code), data=females_no_outliers, REML = FALSE)


#view models
summary(F_nose_wing_5KM_no_outliers)

```

##Generate Plots 
Plot relationship between % impervious surface (5 km buffer) and body length for females and males. 
```{r}

# Create a scatterplot for Females
p_females <- ggplot(females, aes(x = scaled_5KM, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +  # Add the points
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), color = "red") +  # Add the regression line (quadratic)
  labs(
    x = "% Impervious Surface (scaled)",
    y = "Body Length (mm)",
    title = "Females"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18),  # Center and increase title size
    panel.grid = element_blank(),  # Remove grid lines
    axis.text.x = element_text(size = 14),  # Increase x-axis tick label size
    axis.text.y = element_text(size = 14),  # Increase y-axis tick label size
    axis.title.x = element_text(size = 16),  # Increase x-axis title size
    axis.title.y = element_text(size = 16),  # Increase y-axis title size
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White plot background
  )

# Save the plot for Females
ggsave("females_body_size_600dpi.png", plot = p_females, dpi = 600, width = 8, height = 6, units = "in")

# Create a scatterplot for Males
p_males <- ggplot(males, aes(x = scaled_5KM, y = `Nose-To-Wing-Tip-(mm)`)) +
  geom_point() +  # Add the points
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), color = "blue") +  # Add the regression line (quadratic)
  labs(
    x = "% Impervious Surface (scaled)",
    y = "Body Length (mm)",
    title = "Males"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18),  # Center and increase title size
    panel.grid = element_blank(),  # Remove grid lines
    axis.text.x = element_text(size = 14),  # Increase x-axis tick label size
    axis.text.y = element_text(size = 14),  # Increase y-axis tick label size
    axis.title.x = element_text(size = 16),  # Increase x-axis title size
    axis.title.y = element_text(size = 16),  # Increase y-axis title size
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White plot background
  )

# Save the plot for Males
ggsave("males_body_size_600dpi.png", plot = p_males, dpi = 600, width = 8, height = 6, units = "in")


```
Calculate mean nose-to-wingtip length for most (>50% IP) and least (< 15% IP) urbanized areas for males and females.

```{r}
# Mean for %IP-5KM 50% or more
mean_50_or_more_F <- mean(females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` >= 50], na.rm = TRUE)

# Mean for %IP-5KM 15% or less
mean_15_or_less_F <- mean(females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` <= 15], na.rm = TRUE)

# Print the results
cat("Mean Nose-To-Wing-Tip for %IP-5KM 50% or more:", mean_50_or_more_F, "mm\n")
cat("Mean Nose-To-Wing-Tip for %IP-5KM 15% or less:", mean_15_or_less_F, "mm\n")

# Calculate standard deviation for %IP-5KM 50% or more
stddev_50_or_more_F <- sd(females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` >= 50], na.rm = TRUE)

# Calculate standard deviation for %IP-5KM 15% or less
stddev_15_or_less_F <- sd(females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` <= 15], na.rm = TRUE)

# Perform t-test to compare means of two groups
t_test_result_F <- t.test(
  females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` >= 50],
  females$`Nose-To-Wing-Tip-(mm)`[females$`%IP-5KM` <= 15],
  alternative = "two.sided",  # Two-tailed test
  na.rm = TRUE
)

# Print the standard deviations
cat("Standard deviation for %IP-5KM 50% or more:", stddev_50_or_more_F, "\n")
cat("Standard deviation for %IP-5KM 15% or less:", stddev_15_or_less_F, "\n")

# Print the t-test result
print(t_test_result_F)







# Mean for %IP-5KM 50% or more
mean_50_or_more_M <- mean(males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` >= 50], na.rm = TRUE)

# Mean for %IP-5KM 15% or less
mean_15_or_less_M <- mean(males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` <= 15], na.rm = TRUE)

# Print the results
cat("Mean Nose-To-Wing-Tip for %IP-5KM 50% or more:", mean_50_or_more_M, "mm\n")
cat("Mean Nose-To-Wing-Tip for %IP-5KM 15% or less:", mean_15_or_less_M, "mm\n")

# Calculate standard deviation for %IP-5KM 50% or more
stddev_50_or_more_M <- sd(males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` >= 50], na.rm = TRUE)

# Calculate standard deviation for %IP-5KM 15% or less
stddev_15_or_less_M <- sd(males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` <= 15], na.rm = TRUE)

# Perform t-test to compare means of two groups
t_test_result_M <- t.test(
  males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` >= 50],
  males$`Nose-To-Wing-Tip-(mm)`[males$`%IP-5KM` <= 15],
  alternative = "two.sided",  # Two-tailed test
  na.rm = TRUE
)

# Print the standard deviations
cat("Standard deviation for %IP-5KM 50% or more:", stddev_50_or_more_M, "\n")
cat("Standard deviation for %IP-5KM 15% or less:", stddev_15_or_less_M, "\n")

# Print the t-test result
print(t_test_result_M)


```













#Analysis of Thermal Tolerance Data
Importing Data:
```{r}
h1 <- read_csv("/Users/brennalevine/Desktop/SLF_body_size_ICB_ms/Levine_et_al_heat_tolerance.csv")
```

Split the hot and cold data into male and female data sets.
```{r}
#Separating into male & female data sets
#subset to male data
hotmales <- h1 %>%
  filter(`Sex(M/F)` == 'M') %>%
  select(`ID`, `Length-Nosetoabdomen(mm)`, `Length-NosetoWingtip(mm)`, `AbdomenLength(mm)`, `Weight(g)`, `CTmax(s)`, `Temperature(C)`)

#subset to female data
hotfemales <- h1 %>%
  filter(`Sex(M/F)` == 'F') %>%
  select(`ID`, `Length-Nosetoabdomen(mm)`, `Length-NosetoWingtip(mm)`, `AbdomenLength(mm)`, `Weight(g)`, `CTmax(s)`, `Temperature(C)`)

#log transform the response variable
hotmales$`CTmax(s)LOG` <- log(hotmales$`CTmax(s)`)
hotfemales$`CTmax(s)LOG` <- log(hotfemales$`CTmax(s)`)
```

Removing unreliable data/mismeasurements/data collection issues:
```{r}
#hotfemales
rowsToRemove_hF <- c(2, 3, 4, 8) 
hotfemalesRowsRemoved <- hotfemales[-rowsToRemove_hF, ]

#hotmales
rowsToRemove_hM <- c(2) 
hotmalesRowsRemoved <- hotmales[-rowsToRemove_hM, ]
```


Calculate numbers of males and females per treatment:
```{r}
nrow(hotfemalesRowsRemoved)
nrow(hotmalesRowsRemoved)
```


Compare body length and weight between males and females.
```{r}
# Load your data (replace with your actual file paths)
males <- read.csv("hotmalesRowsRemoved.csv")
females <- read.csv("hotfemalesRowsRemoved.csv")

# T-test for Length-NosetoWingtip
wingtip_ttest <- t.test(hotfemalesRowsRemoved$`Length-NosetoWingtip(mm)`, hotmalesRowsRemoved$`Length-NosetoWingtip(mm)`, var.equal = FALSE)
print(wingtip_ttest)

# T-test for Weight
weight_ttest <- t.test(hotfemalesRowsRemoved$`Weight(g)`, hotmalesRowsRemoved$`Weight(g)`, var.equal = FALSE)
print(weight_ttest)

# Correlation between Length-NosetoWingtip and Weight
male_corr <- cor.test(hotmalesRowsRemoved$`Length-NosetoWingtip(mm)`, hotmalesRowsRemoved$`Weight(g)`)
female_corr <- cor.test(hotfemalesRowsRemoved$`Length-NosetoWingtip(mm)`, hotfemalesRowsRemoved$`Weight(g)`)
print(male_corr)
print(female_corr)

# Assumptions check: Q-Q plots for normality
par(mfrow=c(2, 2)) # Arrange plots in 2x2 grid

# Q-Q plot for male wingtip length
qqPlot(hotmalesRowsRemoved$`Length-NosetoWingtip(mm)`, main="Q-Q Plot: Male Wingtip Length")
# Q-Q plot for female wingtip length
qqPlot(hotfemalesRowsRemoved$`Length-NosetoWingtip(mm)`, main="Q-Q Plot: Female Wingtip Length")

# Plot distributions
ggplot() +
  geom_histogram(data=hotmalesRowsRemoved, aes(x=`Length-NosetoWingtip(mm)`), fill="blue", alpha=0.5, bins=20) +
  geom_histogram(data=hotfemalesRowsRemoved, aes(x=`Length-NosetoWingtip(mm)`), fill="red", alpha=0.5, bins=20) +
  labs(title = "Body Length: Males vs Females", x = "Body Length (mm)", y = "Count")

ggplot() +
  geom_histogram(data=hotmalesRowsRemoved, aes(x=`Weight(g)`), fill="blue", alpha=0.5, bins=20) +
  geom_histogram(data=hotfemalesRowsRemoved, aes(x=`Weight(g)`), fill="red", alpha=0.5, bins=20) +
  labs(title = "Weight: Males vs Females", x = "Weight (g)", y = "Count")

# Scatter plots for correlation
ggplot(hotmalesRowsRemoved, aes(x = `Length-NosetoWingtip(mm)`, y = `Weight(g)`)) +
  geom_point(color = 'blue') +
  labs(title = paste("Male Correlation: r =", round(male_corr$estimate, 2), ", p =", round(male_corr$p.value, 3)),
       x = "Body Length (mm)", y = "Weight (g)")

ggplot(hotfemalesRowsRemoved, aes(x = `Length-NosetoWingtip(mm)`, y = `Weight(g)`)) +
  geom_point(color = 'red') +
  labs(title = paste("Female Correlation: r =", round(female_corr$estimate, 2), ", p =", round(female_corr$p.value, 3)),
       x = "Length-NosetoWingtip (mm)", y = "Weight (g)")

```

```{r}

# Calculate mean and standard deviation for weight and Nose-to-Wingtip length for males
male_summary <- hotmalesRowsRemoved %>%
  summarise(
    Mean_Weight = mean(`Weight(g)`, na.rm = TRUE),
    SD_Weight = sd(`Weight(g)`, na.rm = TRUE),
    Mean_Length = mean(`Length-NosetoWingtip(mm)`, na.rm = TRUE),
    SD_Length = sd(`Length-NosetoWingtip(mm)`, na.rm = TRUE)
  )

# Calculate mean and standard deviation for weight and Nose-to-Wingtip length for females
female_summary <- hotfemalesRowsRemoved %>%
  summarise(
    Mean_Weight = mean(`Weight(g)`, na.rm = TRUE),
    SD_Weight = sd(`Weight(g)`, na.rm = TRUE),
    Mean_Length = mean(`Length-NosetoWingtip(mm)`, na.rm = TRUE),
    SD_Length = sd(`Length-NosetoWingtip(mm)`, na.rm = TRUE)
  )

# Print the summaries
print(male_summary)

print(female_summary)


```


Test effect of body length on CTmax.
```{r}
# Load necessary libraries
library(lmtest)

# Plot for males
male_plot <- ggplot(hotmalesRowsRemoved, aes(x = `Length-NosetoWingtip(mm)`, y = `CTmax(s)`)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Males",
    x = "Body Length (mm)",
    y = expression(paste("Time to CT"["max"], " (s)"))
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Save the male plot as a PNG file with 600 DPI
ggsave("male_heat_tolerance_600dpi.png", plot = male_plot, dpi = 600, width = 8, height = 6, units = "in")

# Plot for females
female_plot <- ggplot(hotfemalesRowsRemoved, aes(x = `Length-NosetoWingtip(mm)`, y = `CTmax(s)`)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Females",
    x = "Body Length (mm)",
    y = expression(paste("Time to CT"["max"], " (s)"))
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Save the female plot as a PNG file with 600 DPI
ggsave("female_heat_tolerance_600dpi.png", plot = female_plot, dpi = 600, width = 8, height = 6, units = "in")

# ---- Model Assumptions ----

# 1. Normality of Residuals: Q-Q plot and Shapiro-Wilk test
par(mfrow = c(2, 2))  # Arrange plots in 2x2 grid

# Q-Q plot for male residuals
qqPlot(male_lm_nose, main = "Q-Q Plot: Male Residuals")
shapiro_test_male_nose <- shapiro.test(residuals(male_lm_nose))
cat("Shapiro-Wilk Test for Male Residuals: W =", shapiro_test_male_nose$statistic, ", p =", shapiro_test_male_nose$p.value, "\n")

# Q-Q plot for female residuals
qqPlot(female_lm_nose, main = "Q-Q Plot: Female Residuals")
shapiro_test_female_nose <- shapiro.test(residuals(female_lm_nose))
cat("Shapiro-Wilk Test for Female Residuals: W =", shapiro_test_female_nose$statistic, ", p =", shapiro_test_female_nose$p.value, "\n")

# 2. Homoscedasticity: Residuals vs Fitted plot and Breusch-Pagan test

# Residuals vs Fitted for males
plot(male_lm_nose, which = 1, main = "Residuals vs Fitted: Male")
bptest_male_nose <- bptest(male_lm_nose)
cat("Breusch-Pagan Test for Male Residuals: p =", bptest_male_nose$p.value, "\n")

# Residuals vs Fitted for females
plot(female_lm_nose, which = 1, main = "Residuals vs Fitted: Female")
bptest_female_nose <- bptest(female_lm_nose)
cat("Breusch-Pagan Test for Female Residuals: p =", bptest_female_nose$p.value, "\n")


```



Test effect of weight on CTmax.
```{r}

# Linear regression for males
male_lm_weight <- lm(`CTmax(s)` ~ `Weight(g)`, data = hotmalesRowsRemoved)

# Linear regression for females
female_lm_weight <- lm(`CTmax(s)` ~ `Weight(g)`, data = hotfemalesRowsRemoved)

# Plot for males with centered title and custom axis title for Time to CTmax (with CTmax as a subscript)
ggplot(hotmalesRowsRemoved, aes(x = `Weight(g)`, y = `CTmax(s)`)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Males", 
       x = "Weight (g)", 
       y = expression(paste("Time to CT", max, "(s)"))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Plot for females with centered title and custom axis title for Time to CTmax (with CTmax as a subscript)
ggplot(hotfemalesRowsRemoved, aes(x = `Weight(g)`, y = `CTmax(s)`)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Females", 
       x = "Weight (g)", 
       y = expression(paste("Time to CT", max, "(s)"))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# ---- Model Assumptions ----

# 1. Normality of Residuals: Q-Q plot and Shapiro-Wilk test
par(mfrow = c(2, 2))  # Arrange plots in 2x2 grid

# Q-Q plot for male residuals
qqPlot(male_lm_weight, main = "Q-Q Plot: Male Residuals")
shapiro_test_male <- shapiro.test(residuals(male_lm_weight))
cat("Shapiro-Wilk Test for Male Residuals: W =", shapiro_test_male$statistic, ", p =", shapiro_test_male$p.value, "\n")

# Q-Q plot for female residuals
qqPlot(female_lm_weight, main = "Q-Q Plot: Female Residuals")
shapiro_test_female <- shapiro.test(residuals(female_lm_weight))
cat("Shapiro-Wilk Test for Female Residuals: W =", shapiro_test_female$statistic, ", p =", shapiro_test_female$p.value, "\n")

# 2. Homoscedasticity: Residuals vs Fitted plot and Breusch-Pagan test

# Residuals vs Fitted for males
plot(male_lm_weight, which = 1, main = "Residuals vs Fitted: Male")
bptest_male <- bptest(male_lm_weight)
cat("Breusch-Pagan Test for Male Residuals: p =", bptest_male$p.value, "\n")

# Residuals vs Fitted for females
plot(female_lm_weight, which = 1, main = "Residuals vs Fitted: Female")
bptest_female <- bptest(female_lm_weight)
cat("Breusch-Pagan Test for Female Residuals: p =", bptest_female$p.value, "\n")


```









